---
title: "project_data_sets"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
```{r} 
library(tidyverse)
library(ggplot2)
```
## Links for data https://open.toronto.ca/catalogue/?vocab_civic_issues=Poverty%20reduction&n=1&sort=score%20desc
```{r}
callerData=read_csv("data/211_CallerData.csv")
shelter_data=read_csv("data/14100353.csv")
shelter_2019 = read_csv('data/shelter2019.csv')

tor_data=read_csv('data/TorontoMeasureData.csv')

```

```{r}
callerData%>%filter(Level1Name=="Basic Needs")%>% 
  filter(NeedWasMet==FALSE)%>%group_by(Level3Name)
  
```


finds the occuapncy rate of variious sheleters across the city 
```{r}
full_shelter=shelter_2019%>%filter(OCCUPANCY==CAPACITY)%>%
  group_by(OCCUPANCY_DATE,SHELTER_NAME)%>%distinct(OCCUPANCY_DATE,SHELTER_NAME,.keep_all = TRUE)%>%ungroup()%>%
  group_by(SHELTER_NAME,SHELTER_POSTAL_CODE)%>%summarise(days_full=n())%>%arrange(desc(days_full))

full_shelter

full_shelter%>%ungroup()%>%summarise(Avg_days_full=(mean(days_full)))

```
visualiztion of previous graph(will fix later)
```{r}
full_shelter%>%ggplot(aes(x=SHELTER_POSTAL_CODE,y=days_full,fill=SHELTER_POSTAL_CODE),size=y)+geom_bar(stat="identity")+ coord_flip()
```


SOcial housing demand for years 2017+ seperated quartely 
```{r}

tor_data%>%filter(Year>=2017)%>%
  filter(str_detect(str_to_lower(`Measure Name`),pattern="size"))%>%group_by(Year)%>%
  unite(Time,c(Period,Year),sep="-")


tor_data%>%filter(Year>=2017)%>%
  filter(str_detect(str_to_lower(`Measure Name`),pattern="size"))%>%group_by(Year)%>%
  unite(Time,c(Period,Year),sep="-")%>%
  ggplot(aes(x=Time,y=Value,group=1))+geom_line(stat="identity")+labs(title="Wait list Social Housing")

tor_data%>%filter(Year>=2017)%>%
  filter(str_detect(str_to_lower(`Measure Name`),pattern="size"))%>%group_by(Year)%>%
  unite(Time,c(Period,Year),sep="-")%>%
  ggplot(aes(x=Time,y=Value,group=1))+geom_bar(stat="identity")+labs(title="Wait list Social Housing")

```

SOcial housing demand for years by year from 2007+
```{r}

tor_data%>%filter(str_detect(str_to_lower(`Measure Name`),pattern="size"))%>%
  group_by(Year)%>%
  summarise(total_size=sum(Value))


tor_data%>%filter(str_detect(str_to_lower(`Measure Name`),pattern="size"))%>%
  group_by(Year)%>%
  summarise(total_size=sum(Value))%>%
  ggplot(aes(x=Year,y=total_size,group=1))+geom_line()+labs(title="Wait list Social Housing")


tor_data%>%filter(str_detect(str_to_lower(`Measure Name`),pattern="size"))%>%
  group_by(Year)%>%
  summarise(total_size=sum(Value))%>%
  ggplot(aes(x=Year,y=total_size,group=1))+geom_bar(stat="identity")+labs(title="Wait list Social Housing")

```
CONSLUSION: starting is staring to tackle this issue



Looking at food bank use buy individuals for years 2007+ (might be helpful)
```{r}
tor_data%>% filter(str_detect(str_to_lower(`Measure Name`),pattern="food"))%>%
  group_by(Year)%>%summarise(total_usage=sum(Value))
 




tor_data%>% filter(str_detect(str_to_lower(`Measure Name`),pattern="food"))%>%
  group_by(Year)%>%summarise(Total_usage=sum(Value))%>%
  ggplot(aes(x=Year,y=Total_usage,group=1))+geom_line(stat="identity")+labs(title="Food bank Usage")

tor_data%>% filter(str_detect(str_to_lower(`Measure Name`),pattern="food"))%>%
  group_by(Year)%>%summarise(Total_usage=sum(Value))%>%
  ggplot(aes(x=Year,y=Total_usage,group=1))+geom_bar(stat="identity")+labs(title="Food bank Usage")

```

looking at shelter use by individuals for years 2017+(looked at quarterly)
```{r}
tor_data%>%filter(Year>=2017)%>%
  filter(str_detect(str_to_lower(`Measure Name`),pattern="singles"))%>%group_by(Year)%>%
  unite(Time,c(Period,Year),sep="-")

tor_data%>%filter(Year>=2017)%>%
  filter(str_detect(str_to_lower(`Measure Name`),pattern="size"))%>%group_by(Year)%>%
  unite(Time,c(Period,Year),sep="-")%>%
   ggplot(aes(x=Time,y=Value,group=1))+geom_line(stat="identity")+labs(title="Individuals Shelter Use")

tor_data%>%filter(Year>=2017)%>%
  filter(str_detect(str_to_lower(`Measure Name`),pattern="size"))%>%group_by(Year)%>%
  unite(Time,c(Period,Year),sep="-")%>%
   ggplot(aes(x=Time,y=Value,group=1))+geom_bar(stat="identity")+labs(title="Individuals Shelter Use")


```

looking at shelter use by individuals for years 2007(looked at anually)
```{r}
tor_data%>%filter(str_detect(str_to_lower(`Measure Name`),pattern="singles"))%>%
  group_by(Year)%>%
  summarise(total_usage=sum(Value))
  
tor_data%>%filter(str_detect(str_to_lower(`Measure Name`),pattern="singles"))%>%
  group_by(Year)%>%
  summarise(total_usage=sum(Value))%>%
   ggplot(aes(x=Year,y=total_usage,group=1))+geom_line(stat="identity")+labs(title="Individuals Shelter Use")


tor_data%>%filter(str_detect(str_to_lower(`Measure Name`),pattern="singles"))%>%
group_by(Year)%>%
summarise(total_usage=sum(Value))%>%
ggplot(aes(x=Year,y=total_usage,group=1))+geom_bar(stat="identity")+labs(title="Individuals Shelter Use")

```
USEFUL DATA FOR OUR TOPIC


looking at shelter use by FAmilies for years 2017(looked at quarterly)(can look at oher years if needed to dectef if there is the same pattern)
```{r}

tor_data%>%filter(Year>=2017)%>%
  filter(str_detect(str_to_lower(`Measure Name`),pattern="families"))%>%group_by(Year)%>%
  unite(Time,c(Period,Year),sep="-")


tor_data%>%filter(Year>=2017)%>%
  filter(str_detect(str_to_lower(`Measure Name`),pattern="families"))%>%group_by(Year)%>%
  unite(Time,c(Period,Year),sep="-")%>%
   ggplot(aes(x=Time,y=Value,group=1))+geom_line(stat="identity")+labs(title="Family Shelter Use")


tor_data%>%filter(Year>=2017)%>%
  filter(str_detect(str_to_lower(`Measure Name`),pattern="families"))%>%group_by(Year)%>%
  unite(Time,c(Period,Year),sep="-")%>%
   ggplot(aes(x=Time,y=Value,group=1))+geom_bar(stat="identity")+labs(title="Family Shelter Use")

```

looking at shelter use by FAMILIES for years 2007(looked at anually)
```{r}

tor_data%>%filter(str_detect(str_to_lower(`Measure Name`),pattern="families"))%>%
  group_by(Year)%>%
  summarise(total_usage=sum(Value))


tor_data%>%filter(str_detect(str_to_lower(`Measure Name`),pattern="families"))%>%
  group_by(Year)%>%
  summarise(total_usage=sum(Value))%>%
   ggplot(aes(x=Year,y=total_usage,group=1))+geom_line(stat="identity")+labs(title="Family Shelter Use")


tor_data%>%filter(str_detect(str_to_lower(`Measure Name`),pattern="families"))%>%
group_by(Year)%>%
summarise(total_usage=sum(Value))%>%
ggplot(aes(x=Year,y=total_usage,group=1))+geom_bar(stat="identity")+labs(title="Family Shelter Use")


```


WILL want futher investigation as why the dramatic increase in the last 2 yrs(my guess is the rising house market, and wages have not kept up with the inflation rate(will need to confirm later))